MS797 Glossary

### 2022-02-05 _academics/ms797-glossary/ms797-glossary.Rmd
typo
Many additions

---
title: "MS797 Glossary"
description: |
  Glossary for MS797 coursework
author:
  - name: Steve Linberg
    url: {}
date: 2022-01-30
output:
 distill::distill_article:
   self_contained: false
   toc: true
# bibliography: PS753.bib
---

@@

<p style="background: #eee">
Academics &gt; [MS 797](../ms797/index.html)
&gt; Glossary
</p>

Page numbers in parenthesis after terms, from [ISLR](https://www.statlearning.com/) 2nd edition.

@@

total sum of squares (TSS) (70)
: the total variance in the response $Y$
$$\text{TSS} = \sum(y_i - \bar{y})^2$$
compare with RSS, the amount of variability left unexplained after the regression. TSS - RSS is the amount of variability (or error) explained by the regression.

@@

correlation (70)
: a measure of the linearity of the relationship between $X$ and $Y$; values close to 0 indicate weak-to-no relationship, values near 1 or -1 indicate strong positive or negative correlation
(horrible formula on p.70)

>>>>

---
title: "MS797 Glossary"
description: |
  Glossary for MS797 coursework
author:
  - name: Steve Linberg
    url: {}
date: 2022-01-30
output:
 distill::distill_article:
   self_contained: false
   toc: true
---

@@

<p style="background: #eee">
Academics &gt; [MS797](../ms797/index.html)
&gt; Glossary
</p>

Page numbers in parenthesis after terms, from [ISLR](https://www.statlearning.com/) 2nd edition. Non-page numbers indicate other sources; "biostats" references material from Biostatistics 690Z (Health Data Science: Statistical Modeling), fall 2021.

@@

model sum of squares (MSS) (biostats)
: Also sometimes **ESS**, "explained sum of squares": the total variance in the response $Y$ that can be accounted for by the model
$$\text{MSS} = \sum(\hat{y_i} - \bar{y})^2$$

residual sum of squares (RSS) (biostats)
: the total variance in the response $Y$ that cannot be accounted for by the model
$$\text{RSS} = \sum(y_i - \hat{y_i})^2$$
also
$$\text{RSS} = e_i^2 + e_2^2 + \dots + e_n^2$$
or
$$\text{RSS} = (y_1 - \hat{\beta_0} - {\hat{\beta_1}x_1})^2 + (y_2 - \hat{\beta_0} - {\hat{\beta_1}x_2})^ + \dots + (y_n - \hat{\beta_0} - {\hat{\beta_1}x_n})^2$$

total sum of squares (TSS) (70)
: the total variance in the response $Y$; the total variability of the response about its mean
$$\text{TSS} = \sum(y_i - \bar{y})^2$$
compare with RSS, the amount of variability left unexplained after the regression. TSS - RSS is the amount of variability (or error) explained by the regression (MSS).

**NOTE**: there is a nice visual [here on stackexchange](https://stats.stackexchange.com/a/512216/338749); if anybody knows how to tell Zotero to use a custom bibtex citation entry over the ones it generates, please let me know so I can integrate it better here :frown:

@@

correlation (70)
: a measure of the linearity of the relationship between $X$ and $Y$; values close to 0 indicate weak-to-no relationship, values near 1 or -1 indicate strong positive or negative correlation
(horrible formula on p.70 I don't have time to LaTeX at the moment)

standard linear regression model (72)
: The modal used for standard linear models, used to interpret the the effect on $Y$ of a one-unit increase in any predictor $\beta_j$ while holding all other predictors constant
$$Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \dots + \beta_pX_p + \epsilon$$

variable selection (78)
: the task of refining a model to include only the variables associated with the response

null model (79)
: a model that conatins an intercept, but no predictors; used as a first stage in forward selection

forward selection (79)
: a variable selection method that starts with a null model, and then runs simple linear regressions on all predictors $p$ and adding the one that results in the lowest RSS; repeated until some threshold is reached

backwards selection (79)
: a variable selection method that starts with a model with all predictors, and removing the one with the lowest $p$-value until all remaining predictors are significant, whether by $p$-value or some other criterion

mixed selection (79)
: a hybrid approach starting with a null model, adding predictors one at a time that produce the best fit, and removing any that acquire a larger $p$-value in the process until all predictors are added or eliminated

interaction (81)
: when predictors affect each other, in addition to providing their own effect on the model

confidence interval (82)
: a range with a percentage component in which there is that percentage chance that the true value of an estimated parameter lies; a 95% confidence interval is a range in which we can be 95% certain $f(X)$ will be found

prediction interval (82)
: similar to confidence interval, but a prediction range within which we are $X$% certain that any singular future observation will fall, rather than a statistic like an overall mean; a 95% prediction interval is a range in which we are confident that 95% of future observations will fall. Prediction ranges are substantially wider than confidence intervals.

qualitative predictor / factor (83)
: a categorical predictor with a fixed number of factors, like "yes" / "no" or "red" / "yellow" / "green"

dummy variable (83)
: a numeric representation of a factor to use in a model, as in representing "yes" / "no" factor variables as 1 / 0 in a regression

baseline (86)
: the factor level where there is no dummy variable; a factor with 3 levels will use 2 dummy variables, with the factor's absence signifying the 3rd value (usually 0)

additivity assumption (87)
: the assumption that the association between a predictor $X$ and the response $Y$ does not depend on the value of other predictors; used by the standard linear regression model

linearity assumption (87)
: the assumption, also used by the standard linear regression model, that unit changes in $X_j$ result in the same change to Y regardless of its value

interaction term (88)
: the product of two predictors in a multiple regression model, quantifying their effect on each other

main effect (89)
: isolated effects; the effect of a single predictor on the outcome

hierarchical principle (89)
: the principle that main effects should be left in a model even if they are statistically insignificant, if they are also part of an interaction that is significant

polynomial regression (91)
: an extension of linear regression to accommodate non-linear relationships

residual plot (93)
: a plot of the residuals or errors ($e_i = y_i - \hat{y}_i$), used to check for non-linearity (a potential problem that would likely indicate something was missed in the model)

time series (94)
: data consisting of observation made at discrete points in time

tracking (95)
: when (residuals / variables?) tend to have similar values

heteroscedasticity (96)
: non-constant variances in errors; "unequal scatter"

homoscedasticity (extra)
: constant variances in errors; follows the assumption of equal variance required by most methods

weighted least squares (97)
: an extension to ordinary least squares used in circumstances of heteroscedasticity, to weight data points proportionally with the inverse variances

outlier (97)
: an observation whose value is very far from its predicted value

studentized residual (98)
: a residual divided by its estimated standard error; observations with student errors higher than 3 (indicating 3 standard deviations) are likely outliers

high leverage (98)
: observations with an unusual $x_i$ value, far from other / expected $x$ values

leverage statistic (99)
: a quantification of a point's leverage
$$h_i = \frac{1}{n} + \frac{(x_i - \bar{x})^2}{\sum_{i'=1}^{n}(x_{i`} - \bar{x})^2}$$

collinearity (99)
: when two or more predictor variables are closely related to each other

power (101)
: the probability of a test correctly detecting a nonzero coefficient (and correctly rejecting $H_0 : \beta_j = 0$)

multicollinearity (102)
: when collinearity exists between three or more predictors even when no pair of predictors is collinear (or correlated)

variance inflation factor (102)
: "the ratio of the variance of $\hat{\beta_j}$ when fitting the full model divided by the variance of $\hat{\beta_j}$ on its own"; smallest possible value of 1 indicates the absence of collinearity, 5-10 indicates a "problematic amount". Hairy forumla not reproduced.
